---
title: "Analyze Play-by-Play NBA Data (Sample)"
author: "kyle wurtz"
date: "September 21, 2016"
output: 
  html_document:
    toc: true
    theme: yeti
    code_folding: show
---

## Overview
This is a sample analysis to test the viability of using play-by-play NBA data from [BigDataBall.com](https://www.bigdataball.com/nba-play-by-play-logs) for analysis. I've been looking for a source of play-by-play data that would allow me to calculate things like effective field goal percentage (eFG%) of teammates while a player is on vs. off the court. The play-by-play data that's available at places like Basketball-Reference.com and NBA.com don't have the necessary detail (e.g., they don't describe who is on the court for a given play), but BigDataBall appears to. The data comes with a cost, though, so I wanted to run a sample analysis with their sample file to see if I liked the format and was willing to purchase full seasons of the data.

Given that broader objective, this analysis calculates the relative eFG% of teammates while each player is on vs. off the court for both teams in Game 4 of the 2016 NBA finals.

## Setup
### Load Packages
```{r load packages, warning = FALSE, message = FALSE}
require(tidyverse)
require(magrittr)
require(rvest)
require(ggthemes)
require(pander)
```

### Custom Functions
```{r filter data frame to on court}
filter_on = function(df, nm = "name", h_o_a = "home") {
  if (h_o_a == "home") {
    df %>% 
      filter(
        h1 == nm | h2 == nm | h3 == nm | h4 == nm | h5 == nm
      )
  } else if (h_o_a == "away") {
    df %>% 
      filter(
        a1 == nm | a2 == nm | a3 == nm | a4 == nm | a5 == nm
      )
  } else {
    "Error! Select 'home' or 'away' for h_o_a."
  }
}
```

```{r filter data frame to off court}
filter_off = function(df, nm = "name", h_o_a = "home") {
  if (h_o_a == "home") {
    df %>% 
      filter(
        h1 != nm, h2 != nm, h3 != nm, h4 != nm, h5 != nm
      ) 
  } else if (h_o_a == "away") {
    df %>% 
      filter(
        a1 != nm, a2 != nm, a3 != nm, a4 != nm, a5 != nm
      )
  } else {
    "Error! Select 'home' or 'away' for h_o_a."
  }
}
```

```{r filter to plays for eFG calculation}
# filters to just shots by the same team as the player by someone other than the player
filter_eFG = function(df, nm = "name", tm = "teamAbbrev")  {
  df %>% 
    filter(
      team == tm,
      event_type %in% c("shot", "miss"), # want shots w/o free throws
      player != nm # want shots from teammates
    )
}
```

```{r calculate eFG}
eFG = function(df, made_col = "event_type", made_text = "shot", points_col = "points") {
  (sum(df[made_col] == made_text) + 0.5 * sum(df[points_col] == 3)) / nrow(df)
}
```


```{r find ratio of eFG while on to eFG while off and shots for each}
eFGRatio = function(df, nm = "name", tm = "teamAbbrev", h_o_a = "home") {
  df_on = try(df %>% filter_on(nm, h_o_a) %>% filter_eFG(nm, tm))
  df_off = try(df %>% filter_off(nm, h_o_a) %>% filter_eFG(nm, tm))
  
  if (nrow(df_on) == 0) {
    # no plays that meet criteria -> return NA
    eFG_on = NA
    shots_on = NA
  } else {
    eFG_on = eFG(df_on)
    shots_on = nrow(df_on)
  }
  
  if(nrow(df_off) == 0) {
    # no plays that meet criteria -> return NA
    eFG_off = NA
    shots_off = NA
  } else {
    eFG_off = eFG(df_off)
    shots_off = nrow(df_off)
  }
  
  eFG_Ratio = eFG_on / eFG_off
  
  output = list(
    eFG_Ratio = eFG_Ratio, # ratio of eFG of teammates while on vs. while off court
    eFG_on = eFG_on, # eFG of teammates while player is on court
    eFG_off = eFG_off, # eFG of teammates while player is off court
    shots_on = shots_on, # number of shots by teammates while player is on court
    shots_off = shots_off # number of shots by teammates while player is off court
  )
  
  return(output)
}
```


### Inputs
```{r file paths}
location_of_data = "/Users/kylewurtz/Dropbox/R/NBA Play-By-Play/Data/Sample BigDataBall"
data_file_name = "Sample_BigDataBall.csv"
```

```{r read in data}
df = read_csv(file.path(location_of_data, data_file_name))
```


## Work
### Initial Investigation
```{r initial investigation}
glimpse(df)
```

### Effective Field Goal Percentage When CLE Players are On Vs. Off the Court
```{r cle create data frame for output}
cle_players = df %>% 
  select(h1:h5) %>% 
  gather(pos, player) %>% 
  select(player) %>% 
  unique() %>% 
  arrange(player) %>% 
  .[[1]]

cle_eFGs = tibble(
  player = cle_players,
  eFG_Ratio = NA, eFG_on = NA,  eFG_off = NA,
  shots_on = NA, shots_off = NA
)
```

```{r cle calculate eFG while on court, fig.width = 10.5, fig.height = 7, out.width="900px", out.height = "600px"}
for (ply in cle_eFGs$player) {
  print(ply)
  eFGRatio_output = eFGRatio(df, ply, "CLE", "home")
  cle_eFGs$eFG_Ratio[cle_eFGs$player == ply] = eFGRatio_output$eFG_Ratio
  cle_eFGs$eFG_on[cle_eFGs$player == ply] = eFGRatio_output$eFG_on
  cle_eFGs$eFG_off[cle_eFGs$player == ply] = eFGRatio_output$eFG_off
  cle_eFGs$shots_on[cle_eFGs$player == ply] = eFGRatio_output$shots_on
  cle_eFGs$shots_off[cle_eFGs$player == ply] = eFGRatio_output$shots_off
}
cle_eFGs %>% 
  arrange(desc(eFG_Ratio)) %>% 
  mutate(player = factor(player, levels = .[["player"]]),
         ratio_cred = sqrt(pmin(shots_on, shots_off) / pmax(shots_on, shots_off))) %>% 
  ggplot(., aes(x = player, y = eFG_Ratio, fill = ratio_cred)) +
  geom_bar(stat = "identity") + 
  theme_fivethirtyeight() +
  scale_fill_continuous("Credibility") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Ratio of eFG% of Teammates While On Court vs. Off Court")


```

### Effective Field Goal Percentage When GSW Players are On Vs. Off the Court
```{r gsw create data frame for output}
gsw_players = df %>% 
  select(a1:a5) %>% 
  gather(pos, player) %>% 
  select(player) %>% 
  unique() %>% 
  arrange(player) %>% 
  .[[1]]

gsw_eFGs = tibble(
  player = gsw_players,
  eFG_Ratio = NA, eFG_on = NA,  eFG_off = NA,
  shots_on = NA, shots_off = NA
)
```

```{r gsw calculate eFG while on court, fig.width = 10.5, fig.height = 7, out.width="900px", out.height = "600px"}
for (ply in gsw_eFGs$player) {
  eFGRatio_output = eFGRatio(df, ply, "GSW", "away")
  gsw_eFGs$eFG_Ratio[gsw_eFGs$player == ply] = eFGRatio_output$eFG_Ratio
  gsw_eFGs$eFG_on[gsw_eFGs$player == ply] = eFGRatio_output$eFG_on
  gsw_eFGs$eFG_off[gsw_eFGs$player == ply] = eFGRatio_output$eFG_off
  gsw_eFGs$shots_on[gsw_eFGs$player == ply] = eFGRatio_output$shots_on
  gsw_eFGs$shots_off[gsw_eFGs$player == ply] = eFGRatio_output$shots_off
}
gsw_eFGs %>% 
  arrange(desc(eFG_Ratio)) %>% 
  mutate(player = factor(player, levels = .[["player"]]),
         ratio_cred = sqrt(pmin(shots_on, shots_off) / pmax(shots_on, shots_off))) %>% 
  ggplot(., aes(x = player, y = eFG_Ratio, fill = ratio_cred)) +
  geom_bar(stat = "identity") + 
  theme_fivethirtyeight() +
  scale_fill_continuous("Credibility") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Ratio of eFG% of Teammates While On Court vs. Off Court")


```


```{r show credibility }
data_frame(
  lower = 1:100,
  higher = 100,
  ratio = lower / higher,
  sq_ratio = sqrt(lower / higher)
) %>% 
  select(-higher) %>% 
  gather(type, value, -lower) %>% 
  ggplot(., aes(x = lower, y = value, group = type, color = type)) +
    geom_line() +
  theme_fivethirtyeight()
```


